### 2024-03-05 tex/mech-related.tex
[ correction ] first pass over chapter 2

Another main difference is that Autosubst is presented to the user largely as
a black-box implementation of substitution and related lemmas, in contrast to
\citeauthor{AACMM21}'s work exposing the \AgdaRecord{Semantics} bundle to the
user, and having substitution be just one instance.
\citet{ACMM17} and \citet{AACMM21} provide many examples of traversals over
syntax using the same generic environment management as used by substitution.
However, the focus on substitution in Autosubst has meant that reasoning about
substitutions has been given more developed support.
For example, the library provides a tactic \texttt{autosubst} which automates
many equational proofs involving substitutions based on the $\sigma$-calculus of
\citet{ACCL91}.

@@

\citet{Autosubst18} propose some modifications to Autosubst which, as far as I
can tell, have not yet been incorporated, although all of the case studies in
the paper are mechanised in Coq.
In the paper, they adapt and extend the work of \citeauthor{ACMM17} on generic
semantic traversals to cover a variant of system F.
They use the term \emph{multivariate traversal} for the generalisation of
heterogeneous substitution to semantic traversals.
It appears that this work could be followed through to produce syntax
descriptions covering polymorphic calculi, which would provide a route for this
work to be incorporated into the Autosubst library.

@@

Marcelo Fiore and various collaborators have a long line of work aiming for a
categorical account of variable-binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
A recent, particularly relevant paper from this line of work is that of
\citet{FS22}, which mechanises some of this work to obtain a framework similar
in scope to the work of \citet{AACMM21}.
However, there are several differences between the resulting mechanised
frameworks, of methodological, mathematical, and technical nature.

@@

In terms of the underlying theory, both works extend multi-sorted universal
algebra%\todo{cite}
with variable-binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires an explicit substitution
whenever it is used.

>>>>

Another main difference is that Autosubst is presented to the user largely as
a black-box implementation of substitution and related lemmas, in contrast to
\citeauthor{AACMM21}'s work exposing the \AgdaRecord{Semantics} bundle to the
user, and having substitution be just one instance.
\Citet{ACMM17} and \citet{AACMM21} provide many examples of traversals over
syntax using the same generic environment management as used by substitution.
However, the focus on substitution in Autosubst has meant that reasoning about
substitutions has been given more developed support.
For example, the library provides a tactic \texttt{autosubst} which automates
many equational proofs involving substitutions based on the $\sigma$-calculus of
\citet{ACCL91}.
This difference in exposure of the internals can be traced back to the differing
implementation approaches of the two tools.
Autosubst is implemented via metaprogramming in Coq's Ltac sublanguage, whereas
\citeauthor{AACMM21}'s tool is implemented via generic programming within the
main term language of Agda.

@@

\Citet{Autosubst19} describe a new version of the tool --- \emph{Autosubst 2}.
Autosubst 2 continues the Autosubst implementation style as metaprogramming over
Coq, but this time implements the metaprogramming as an external Haskell tool
rather than a collection of Ltac tactics.
Autosubst 2 moves closer to the work of \citeauthor{AACMM21} by producing types
of well scoped terms, with scope-respecting operations over them.
It also adopts the semantic traversals of \citet{ACMM17}, as described by
\citet{Autosubst18}.
Autosubst 2 surpasses both the original Autosubst and the work of
\citet{AACMM21} in its dedicated support for mutually inductive syntax
definitions, which would have to be encoded for use in the other two tools.
\Citet{Autosubst19} give a running example of call-by-value System F,
demonstrating both the handling of polymorphic calculi and mutually inductively
defined syntax --- with expressions and values being mutually defined.
It would be an interesting experiment to port this example to
\citeauthor{AACMM21}'s framework, to check that these features can indeed be
encoded.
It is also conceivable that the syntax descriptions of \citeauthor{AACMM21}
could be extended with explicit support for multiple mutually defined term
sorts.

@@

Marcelo Fiore and various collaborators have a long line of work aiming for a
categorical account of variable binding~\citep{FPT99,Fiore08,FH13,FH10,FM10}.
A recent, particularly relevant paper from this line of work is that of
\citet{FS22}, which mechanises some of this work to obtain a framework similar
in scope to the work of \citet{AACMM21}.
However, there are several differences between the resulting mechanised
frameworks, of methodological, mathematical, and technical nature.

@@

In terms of the underlying theory, both works extend multi-sorted universal
algebra with variable binding.
However, the two extensions are subtly different.
Universal algebra already has a notion of variable, which supports renaming and
substitution, and which allows a given term to be evaluated when the free
variables of that term are assigned semantic values.
\citeauthor{AACMM21} reuse this notion of variable, and allow binding of such
variables in terms.
On the other hand, \citeauthor{FS22} recast the existing variables as
\emph{metavariables}, and introduce a new notion of (bindable) variables
separately into the syntax.
The resulting metavariables can then stand for arbitrary \emph{open} terms, and
thus each one remembers its context and requires an explicit substitution
whenever it is used.

### 2024-03-05 tex/simple.tex
[ writing ] ready for submission
[ correction ] first pass over chapter 2

In this chapter, I review and justify the family of approaches usually used to
represent simple type systems inside dependently typed proof assistants.
These approaches were first presented by \citet{AR99}, who showed a way of
representing well \emph{scoped} terms in a language with polymorphic recursion,
and extended the representation to well \emph{typed} terms in a language with
dependent types.
The representation of terms relies on indexing on both a context --- giving the
types of all the free variables --- and a type for the term itself.
A basic operation on terms is \emph{simultaneous substitution}, which replaces
each variable in the context by a term in another context.
%This simultaneous substitution operation fits the form of composition in a
%multicategory, giving us a connection to categorical logic~\citep{}.
%Specifically, in the case of a simple intuitionistic type system, this is the
%composition of a Cartesian multicategory, which allows all of the terms being
%substituted in to share a context.

This chapter almost entirely reviews prior work.
First, in \cref{sec:agda-primer}, I give an introduction to Agda, the proof
assistant I work in throughout this thesis.
Then, in \cref{sec:terms}, I use a presentation of the dependently typed
encoding of the simply typed $\lambda$-calculus from \citet{AR99} to set
notational conventions.
\Cref{sec:kits} presents the unpublished work of \citet{McBride05},
successively deriving simultaneous renaming and simultaneous substitution for
the terms defined in \cref{sec:terms}.
The rest of the chapter generalises the shared core of renaming and substitution
in two dimensions: in \cref{sec:gen-sem} following \citet{ACMM17} to cover
semantic traversals, and in \cref{sec:gen-syn} following \citet{AACMM21} to
cover a whole range of simply typed syntaxes with binding, rather than just
a specific syntax.
Finally, I review some related work in \cref{sec:mech-related}.

>>>>

In this chapter, I review and justify the family of approaches usually used to
represent simple type systems inside dependently typed proof assistants:
intrinsically well scoped and well typed representations.
The line of work leading to intrinsically well typed representations of syntax
starts with \citet{BH94}, who defined, in Standard ML, a type of
$\lambda$-calculus terms indexed on the set/type of variables they contain, and
gave some operations on them.
Later, \citet{AR99} (working from a suggestion by Hook) and \citet{BP99DeBruijn}
used Haskell's support for nested
datatypes and polymorphic recursion over them to define types of terms indexed
only over \emph{free} variables, with a $\lambda$-abstraction having one fewer
free variable than its immediate subterm.
\Citet{AR99} also showed how to extend this well \emph{scoped} representation of
terms to a well \emph{typed} representation, given a metalanguage with dependent
types.
The representation of terms relies on indexing on both a context --- giving the
types of all the free variables --- and a type for the term itself.
A basic operation on terms is \emph{simultaneous substitution}, which replaces
each variable in the context by a term in another context.
%This simultaneous substitution operation fits the form of composition in a
%multicategory, giving us a connection to categorical logic~\citep{}.
%Specifically, in the case of a simple intuitionistic type system, this is the
%composition of a Cartesian multicategory, which allows all of the terms being
%substituted in to share a context.

This chapter reviews prior work, except for a mildly novel connecting stage at
the start of \cref{sec:gen-sem} introducing the
\AgdaFunction{$\bigcirc$}-modality.
First, in \cref{sec:agda-primer}, I give an introduction to Agda, the proof
assistant I work in throughout this thesis.
Then, in \cref{sec:terms}, I use a presentation of the dependently typed
encoding of the simply typed $\lambda$-calculus from \citet{AR99} to set
notational conventions.
\Cref{sec:kits} presents the work of \citet{BHKM12},
successively deriving simultaneous renaming and simultaneous substitution for
the terms defined in \cref{sec:terms} from a single traversal of the syntax.
The rest of the chapter generalises the shared core of renaming and substitution
in two dimensions: in \cref{sec:gen-sem} following \citet{ACMM17} to cover
semantic traversals, and in \cref{sec:gen-syn} following \citet{AACMM21} to
cover a whole range of simply typed syntaxes with binding, rather than just
a specific syntax.
Finally, I review some related work in \cref{sec:mech-related}.

The Agda code displayed in this chapter is available for further study at
\url{https://github.com/laMudri/thesis/tree/master/agda}, except for the code
in \cref{sec:gen-syn}, which is available at
\url{https://github.com/laMudri/generic-lr/tree/thesis/src/Generic/Simple}.
